Summary of the things that I learned

### 2024-10-08 700-formalization.tex
Address a lot of feedback, do a spellcheck

\section{Equality, Isomorphisms and Equivalences}
One of the lessons during this project was that for formalizing, it is always important to choose te right equality for the job. For example, for two categories $ C $ and $ C^\prime $, one can aim to prove either that $ C = C^\prime $ or that we have a functor $ F : C \to C^\prime $ that has a right and left inverse, or that we have a functor $ F : C \to C^\prime $ that is an adjoint equivalence. Even though for univalent categories, these three notions all coincide, in practice it is really hard to show equality of categories directly: To show $ C = C^\prime $, we would need to show that we have an equality $ H : C_0 = C_0^\prime $, and then that for all $ c, c^\prime : C $
\[ (C(\mathrm{transport}_H(c), \mathrm{transport}_H(c^\prime))) = C^\prime(c, c^\prime). \]
It is usually much easier to show that there exists a functor $ F $ with an inverse, because then we don't have to show these equalities of types, but can suffice with maps between the types and equalities of the objects and morphisms that are mapped. However, for the morphisms we still have transports.

The easiest option is usually to show that we have an adjoint equivalence. That way, we don't have to show equality of objects, but only isomorphism (even though in a univalent category, that is of course the same). Then we only have to show equality of morphisms, and this is usually very doable.

@@

Even though the approach of first defining the categories of all algebras or presheaves and then taking a fiber of this, is necessary to talk about fibrations, there is a drawback to this approach. Morphisms in any fiber category are the displayed morphsims over $ \id T $. Therefore, naively composing two morphisms in our fiber gives a displayed morphism over $ \id T \cdot \id T $ and we need to transport over the equality $ \id T \cdot \id T = \id T $ to get morphisms in our fiber category again. So even though we can prove that, as expected, $ (f \cdot g)(a) = g(f(a)) $ for $ f: \Alg T(A, B) $, $ g: \Alg T(B, C) $ and $ a : A $, this is no longer a definitional equality.

@@

Very early on in the project, a choice was made to use the second approach. Since many definitions require extending a mapping to a tuple, and since we don't very often need to extend tuples or do induction on a tuple, this proved to be the right choice.

@@

In this case, somewhat of a compromise was made. The endomorphism theory was indeed defined using a repeated binary product, and the construction of finite powers from the terminal object and binary products\footnote{Unfortunately, defininig the product of a tuple of objects was much harder than defining the finite power of one object. When generalizing the approach to tuples of objects, some terms suddenly do not have the `correct' type, which has something to do with some equalities like $ ((x_1, \dots, x_n) + (y))_{n + 1} = y $ that do not hold definitionally. Even though this specific problem can be solved using transports, the rest of the proof becomes then much harder because the transports get in the way}. On the other hand, the Yoneda lemma for $ L $-presheaves does indeed use the `usual' notion of products $ L^n $ where the sets are tuple types. Luckily, the library contains a proof that two products are isomorphic, so where the definitions clash, we use the isomorphisms to translate between them. Even though this adds friction, in the end it is pretty straightforward.

@@

It would be interesting to see whether parts of these tactics can be generalized. For example, whether it is possible to create an extensible tactic which shows that some type is a homotopy set, or a mere proposition, and prints the coq statements that can replace the tactic call. Or whether it would be possible to generalize \texttt{generate\_refine} to very qukickly generate very fast and precise rewrite statements for large and complicated goals.

>>>>

\section{Equality, Isomorphisms and Equivalences}
One of the lessons during this project was that for formalizing, it is always important to choose te right equality for the job. For example, for two categories $ C $ and $ C^\prime $, one can aim to prove either that $ C = C^\prime $ or that we have a functor $ F : C \to C^\prime $ that has a right and left inverse, or that we have a functor $ F : C \to C^\prime $ that is an adjoint equivalence. Even though for univalent categories, these three notions all coincide, in practice it is really hard to show equality of categories directly: To show $ C = C^\prime $, we would need to show that we have an equality $ H : C_0 = C_0^\prime $, and then that for all $ c, c^\prime : C $
\[ (C(\mathrm{transport}_H(c), \mathrm{transport}_H(c^\prime))) = C^\prime(c, c^\prime). \]
It is usually much easier to show that there exists a functor $ F $ with an inverse, because then we do not have to show these equalities of types, but can suffice with maps between the types and equalities of the objects and morphisms that are mapped. However, for the morphisms we still have transports.

The easiest option is usually to show that we have an adjoint equivalence. That way, we do not have to show equality of objects, but only isomorphism (even though in a univalent category, that is of course the same). Then we only have to show equality of morphisms, and this is usually very doable.

@@

Even though the approach of first defining the categories of all algebras or presheaves and then taking a fiber of this, is necessary to talk about fibrations, there is a drawback to this approach. Morphisms in any fiber category are the displayed morphisms over $ \id T $. Therefore, naively composing two morphisms in our fiber gives a displayed morphism over $ \id T \cdot \id T $ and we need to transport over the equality $ \id T \cdot \id T = \id T $ to get morphisms in our fiber category again. So even though we can prove that, as expected, $ (f \cdot g)(a) = g(f(a)) $ for $ f: \Alg T(A, B) $, $ g: \Alg T(B, C) $ and $ a : A $, this is no longer a definitional equality.

@@

Very early on in the project, a choice was made to use the second approach. Since many definitions require extending a mapping to a tuple, and since we do not very often need to extend tuples or do induction on a tuple, this proved to be the right choice.

@@

In this case, somewhat of a compromise was made. The endomorphism theory was indeed defined using a repeated binary product, and the construction of finite powers from the terminal object and binary products\footnote{Unfortunately, defining the product of a tuple of objects was much harder than defining the finite power of one object. When generalizing the approach to tuples of objects, some terms suddenly do not have the `correct' type, which has something to do with some equalities like $ ((x_1, \dots, x_n) + (y))_{n + 1} = y $ that do not hold definitionally. Even though this specific problem can be solved using transports, the rest of the proof becomes then much harder because the transports get in the way}. On the other hand, the Yoneda lemma for $ L $-presheaves does indeed use the `usual' notion of products $ L^n $ where the sets are tuple types. Luckily, the library contains a proof that two products are isomorphic, so where the definitions clash, we use the isomorphisms to translate between them. Even though this adds friction, in the end it is pretty straightforward.

@@

It would be interesting to see whether parts of these tactics can be generalized. For example, whether it is possible to create an extensible tactic which shows that some type is a homotopy set, or a mere proposition, and prints the coq statements that can replace the tactic call. Or whether it would be possible to generalize \texttt{generate\_refine} to very quickly generate very fast and precise rewrite statements for large and complicated goals.

### 2024-10-08 920-weak-cartesian-closed.tex
Address a lot of feedback, do a spellcheck

It seems that semi cartesian closed categories were first defined in \cite{hayashi-1985-semifunctors} and then \cite{hoofman-1993-semi-cartesian-closed} gave a slightly different definition, as well as the very related definition of a weak cartesian closed category. The latter paper also contains the construction of the endomorphism $ \lambda $-theory.

\begin{definition}
  Let $ C $ be a category with binary products (we will call the projections $ \pi_1 $ and $ \pi_2 $), and take $ A, B : C $. A \iindex{semi-exponential object} is an object $ A^B : C $, with a morphism $ \mathrm{ev}: C(A^B \times B \to A) $ and a hayashi-1985-semifunctorsfunction
  \[ \mathrm{cur}_X: C(X \times B \to A) \to C(X, A^B) \]
  for all $ X $, such that
  \[ \langle g \cdot \mathrm{cur}_X(f), h \rangle \cdot \mathrm{ev} = \langle g, h \rangle \cdot f \quad \text{and} \quad \mathrm{cur}_Y((g \times \id B) \cdot f) = g \cdot \mathrm{cur}_X(f) \]
  for all $ f: C(X \times B, A) $, $ g : C(Y, X) $ and $ h : C(Y, B) $.
\end{definition}

@@

\begin{remark}
  Since the requirements for semi-exponential objects do not involve the usual universal properties, a weak cartesian closed structure (or a semi-exponential object) is not unique up to isomorphism. For example, see example 2.1 and 2.2 in \cite{hoofman-1993-semi-cartesian-closed} for two distinct weak cartesian closed structures on a category.
\end{remark}

>>>>

It seems that semi cartesian closed categories were first defined in \autocite{hayashi-1985-semifunctors} and then \autocite{hoofman-1993-semi-cartesian-closed} gave a slightly different definition, as well as the very related definition of a weak cartesian closed category. The latter paper also contains the construction of the endomorphism $ \lambda $-theory.

\begin{definition}
  Let $ C $ be a category with binary products (we will call the projections $ \pi_1 $ and $ \pi_2 $), and take $ A, B : C $. A \iindex{semi-exponential object} is an object $ A^B : C $, with a morphism $ \mathrm{ev}: C(A^B \times B \to A) $ and a function
  \[ \mathrm{cur}_X: C(X \times B \to A) \to C(X, A^B) \]
  for all $ X $, such that
  \[ \langle g \cdot \mathrm{cur}_X(f), h \rangle \cdot \mathrm{ev} = \langle g, h \rangle \cdot f \quad \text{and} \quad \mathrm{cur}_Y((g \times \id B) \cdot f) = g \cdot \mathrm{cur}_X(f) \]
  for all $ f: C(X \times B, A) $, $ g : C(Y, X) $ and $ h : C(Y, B) $.
\end{definition}

@@

\begin{remark}
  Since the requirements for semi-exponential objects do not involve the usual universal properties, a weak cartesian closed structure (or a semi-exponential object) is not unique up to isomorphism. For example, see example 2.1 and 2.2 in \autocite{hoofman-1993-semi-cartesian-closed} for two distinct weak cartesian closed structures on a category.
\end{remark}
